% Copyright Page

\thispagestyle{empty}

\vspace*{5cm}

\noindent
{\Large\bfseries Wayland: A Complete Guide to Modern Display Server Architecture}

\vspace{0.5cm}

\noindent
Copyright \textcopyright\ \the\year\ 1ay1

\vspace{0.3cm}

\noindent
\url{https://github.com/1ay1/wayland_book}

\vspace{1cm}

\noindent
\textbf{License}

\vspace{0.5cm}

\noindent
This work is licensed under the Creative Commons Attribution-ShareAlike 4.0 International License. To view a copy of this license, visit:

\url{http://creativecommons.org/licenses/by-sa/4.0/}

\vspace{0.5cm}

\noindent
You are free to:
\begin{itemize}
    \item \textbf{Share} --- copy and redistribute the material in any medium or format
    \item \textbf{Adapt} --- remix, transform, and build upon the material for any purpose, even commercially
\end{itemize}

\noindent
Under the following terms:
\begin{itemize}
    \item \textbf{Attribution} --- You must give appropriate credit, provide a link to the license, and indicate if changes were made.
    \item \textbf{ShareAlike} --- If you remix, transform, or build upon the material, you must distribute your contributions under the same license as the original.
\end{itemize}

\vspace{1cm}

\noindent
\textbf{Disclaimer}

\vspace{0.5cm}

\noindent
The information in this book is provided "as is" without warranty of any kind, either express or implied, including but not limited to the implied warranties of merchantability and fitness for a particular purpose. The author shall not be liable for any damages arising from the use of this information.

\vspace{0.5cm}

\noindent
All trademarks and registered trademarks mentioned in this book are the property of their respective owners. The use of these marks does not imply any affiliation with or endorsement by their owners.

\vspace{1cm}

\noindent
\textbf{Acknowledgments of Open Source Projects}

\vspace{0.5cm}

\noindent
This book discusses many open source projects including Wayland, wlroots, X11, Mesa, and various Wayland compositors. These projects are developed by their respective communities and are subject to their own licenses.

\vspace{1cm}

\noindent
\textbf{Source Repository}

\vspace{0.5cm}

\noindent
The source code for this book is available at:

\url{https://github.com/1ay1/wayland_book}

\vspace{0.5cm}

\noindent
Contributions, corrections, and suggestions are welcome via GitHub issues and pull requests.

\vspace{1cm}

\noindent
First Edition: \today

\noindent
Version 1.0

\clearpage
